---
title: "Mini Data Analysis Milestone 2"
output: github_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome to the rest of your mini data analysis project!

In Milestone 1, you explored your data. and came up with research questions. This time, we will finish up our mini data analysis and obtain results for your data by: 

- Making summary tables and graphs 
- Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

We will also explore more in depth the concept of *tidy data.* 

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

# Instructions

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work here--->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to your mini-analysis GitHub repository, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 50 points: 45 for your analysis, and 5 for overall reproducibility, cleanliness, and coherence of the Github submission. 

**Research Questions**: In Milestone 1, you chose two research questions to focus on. Wherever realistic, your work in this milestone should relate to these research questions whenever we ask for justification behind your work. In the case that some tasks in this milestone don't align well with one of your research questions, feel free to discuss your results in the context of a different research question.

# Learning Objectives

By the end of this milestone, you should:

-   Understand what *tidy* data is, and how to create it using `tidyr`.
-   Generate a reproducible and clear report using R Markdown.
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

# Setup

Begin by loading your data and the tidyverse package below:

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
```

# Task 1: Process and summarize your data 

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (1 point) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:


<!-------------------------- Start your work below ---------------------------->
1. What specific street in Vancouver contains the largest density of trees from different species?

2. What species has the largest height and diameter compared to the species with the smallest height and diameter?

3.Does the street side name have effect on the growth of the trees in terms of height and species, or is it negligible? 

4. Are certain species that are generally tall greatly influenced by the neighbourhood environment they are in or is their growth constant regardless of the condition? 

<!----------------------------------------------------------------------------->

Here, we will investigate your data using various data manipulation and graphing functions.

### 1.2 (8 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

Also make sure that you're using dplyr and ggplot2 rather than base R. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Compute the proportion and counts in each category of one categorical variable across the groups of another categorical variable from your data. Do not use the function `table()`!

**Graphing:**

6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7. Make a graph where it makes sense to customize the alpha transparency.

Using variables and/or tables you made in one of the "Summarizing" tasks: 

8. Create a graph that has at least two geom layers. 
9. Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!


<!------------------------- Start your work below ----------------------------->

#### Analysis Q#1: What specific street in Vancouver contains the largest density of trees from different species?

**Summarizing Q4: Computing Counts and Proportions**
```{r}

#Creating count variable
large_density_count <-vancouver_trees %>%
  group_by(std_street,species_name ) %>%
  summarise(n = n()) 

#Making count variable in descending order
large_density_desc <-arrange(large_density_count, desc(n))

print(large_density_desc)

#Creating proportion variable
large_density_prop <- large_density_desc %>% 
  mutate(n = n / sum(n)) %>% 
  pivot_wider(names_from = std_street, values_from = n)

print(large_density_prop)

#Creating W13AV count variable
W13AV_count <- large_density_count %>%
   filter(std_street %in% c("W 13TH AV"))

print(W13AV_count) 

```

This task has answered my research question #1.In order to determine the specific street in Vancouver with the largest density of a certain type of tree I first created a count variable. This count variable determined the number of species on a certain street, and after I sorted the tibble into descending order to determine the street with the highest count. Thereafter I also determined the proportions of the species as well. From the tables above, it is evident that W 13th Av has the most trees of the Platanoides species with a count of n= 377 and proportion of 0.298. I created a new variable on W13AV_count to determine the relative proportion of Platanoides in comparison to other tree species on this street. 

**Graphing Q8: Creating Graph with Two Geom Layers**
```{r}
#Creating W13AV ggplot with geom layers
W13AV_geom2 <- ggplot(W13AV_count, aes(x = species_name, y = n)) + 
  geom_line(aes(group=1)) + geom_point() + theme_bw()

#Swapping axis sideways and adding title 
W13AV_graph <- W13AV_geom2 + theme(axis.text.x = element_text( size = 5, angle = 90),
          axis.text.y = element_text( size = 5, angle = 45)) 

print(W13AV_graph)

```

In the graph above I incorporated two geom layers of geom_line and geom_point in order to visualize the different species on W13AV using the W13AV_count variable. From this graph I can conclude that Platanoids, Americana and Sylvatica are the top 3 tree species on this street. The difference in the count of Platanoids compared to the other species are tremendous as shown in the graph.  



#### Analysis Q#2: What species has the largest height and diameter compared to the species with the smallest height and diameter?

**Summarizing Q1: Compute range, mean, and two other summary statistics**
```{r}
#Broom Library 
library(broom)

#Determining summary statistics of height and diameter  
stat_species <- vancouver_trees %>% select(height_range_id, diameter)  %>% tidy()

print(stat_species)

#Determining species names with max diameter   
max_diam <- vancouver_trees %>% select(species_name, height_range_id, diameter) %>%  
  filter(diameter == 435)

print(max_diam)

#Determining species names with min diameter   
min_diam <- vancouver_trees %>% select(species_name, height_range_id, diameter) %>% 
  filter(diameter == 0)

print(min_diam)

#Determining species names with max height   
max_height <- vancouver_trees %>% select(species_name, height_range_id, diameter) %>% 
  filter(height_range_id == 10)

print(max_height)

#Determining species names with min height   
min_height <- vancouver_trees %>% select(species_name, height_range_id, diameter) %>% 
    filter( height_range_id == 0)

print(min_height)
```
The task of determining the summary statistics has answered research question #2. I first loaded the broom library in order to complete the analyses. I was able to create a new tibble with summary statistics using the broom package to find the height and diameters of the tree species. By arranging the columns of the Vancouver_trees dataset, I determined that the species with the largest diameter of 435 was Japonica. The other values of maximum height of 10, and minimum height/diameter of 0 does not seem to be significant as there are multiple trees with these dimensions. 

**Graphing Q7: Creating Graph with Alpha Transparency**

```{r}
#Creating dimension_trees ggplot with alpha transparency
dimension_trees <-ggplot(vancouver_trees, aes(x=diameter, y=height_range_id))+ 
  geom_point(size = 2,
            alpha = 0.8) 

print(dimension_trees)

```

In this graph I chose to add in alpha transparency as there were multiple points at the similar heights and diameter ranges which needed to be differentiated from one another. This plot also visualizes the maximum and minimum height ranges and diameters, however species name was not determined. 



#### Analysis Q#3: Does the street side name have effect on the growth of the trees in terms of height and species, or is it negligible? 

**Summarizing Q1: Compute range, mean, and two other summary statistics**
```{r}
#Widening dataset with street side names 
street_wide <- vancouver_trees %>% 
  pivot_wider(names_from = street_side_name,
                values_from = height_range_id)

print(street_wide)

#Creating variable of only street side names 
street_height <- street_wide %>% 
    select(ODD, EVEN, MED, PARK, `BIKE MED`, GREENWAY)

print(street_height)

#Creating summary variable of street side names 
street_height_sum <- tidy(street_height) 

print(street_height_sum)

```
This task partially answered research question 3. For each of the different street side names I was able to determine statistics about the ODD, EVEN, MED, PARK, BIKE MED and GREENWAY locations. This included the mean, standard deviation, median, min, max, range and other values. From the tibble I can determine that the ODD and EVEN street sides have the highest tree height and the ODD sides have the highest mean for height as well. Since the values are simmilar, I can conclude that strret side name does not have an overall big impact on the growth of trees since the number of trees between ODD and EVEN where very similar. However, through this analysis I was unable to answer the question of which specific species is abundant on the street sides. 


**Graphing Q9: Create 3 histograms, with each histogram having different sized bins.**
```{r}
#Creating histogram with bin size 10 
hist_1 <-ggplot(street_height, aes(x = ODD, y = ..density..)) +
        geom_histogram(bins = 10,  fill = "#B3E4F7")

print(hist_1)
#Creating histogram with bin size 20 
hist_2 <-ggplot(street_height, aes(x = ODD, y = ..density..)) +
        geom_histogram(bins = 20,  fill = "#B3E4F7")

print(hist_2)

#Creating histogram with bin size 30 
hist_3 <-ggplot(street_height, aes(x = ODD, y = ..density..)) +
        geom_histogram(bins = 30,  fill = "#B3E4F7")

print(hist_3)
```

Based on the graphs, I can conclude that the third graph with the bin size of 30 is the best layout for the histograms as it completely seperates each column. It is also evident that the ODD street side names have a higher density of trees with a height less than 2.5 and a very low density of trees at a height of 10. This is interesting because although there are trees with a max height of 10, it is not the abundant tree height characteristic. 



#### Analysis Q#4: Are certain species that are generally tall greatly influenced by the neighbourhood environment they are in or is their growth constant regardless of the condition? 

**Summarizing Q2: Compute the number of observations for at least one of your categorical variables**
```{r}
#Determining all trees with max height of 10
height_10 <- vancouver_trees %>% filter(height_range_id == "10")

print(height_10)

#Creating new tibble with neighborhood names 
height_neighbour <- height_10 %>% 
    select(longitude, latitude,neighbourhood_name, height_range_id,species_name)
print(height_neighbour)

#Determining number of observations for each species with max hieght
height_obs <- height_10 %>%
    group_by(species_name) %>%
    summarise(Count = n())

print(height_obs)


```

This analysis helped me answer research quesion #4 as I determined that the Menziesii species has the largest number of trees at the maximum height value of 10. By also looking at the neighborhood names and longtiude and latitudes, it is evident that these trees grow in a close area between Renfrew Collingwood and Riley Park. However, since there are multiple trees with heights of 10 it is unclear which environment/neighborhood is ideal foroptimal growth.  

**Graphing Q8: Creating Graph with Two Geom Layers**

```{r}
#Creating graph of longitude and latitude of Vancouver
max_height_van <- ggplot(vancouver_trees, aes(longitude, latitude)) + 
  geom_point(size = .25, show.legend = FALSE) +
  geom_point(data = height_neighbour, mapping = aes(x = longitude, y = latitude), colour = "red")
  coord_quickmap()
  
print(max_height_van)
```

Based on the height_neighbour variable, I created the graph of Vancouver and highlighted the locations with the trees at a height of 10. The first geom point layer outlines the longitude and latitude of all locations of trees. The second geom layers specifically highlights the largest height trees in the height_neighbour variable.Based on this graph, I can reconfirm that at the longitude at about -123.15 and latitude of 49.24 is where the species of Menziesii can be found. 

<!----------------------------------------------------------------------------->

### 1.3 (2 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->

I am very close to answering my research questions. For question 1, I was able to determine that the specific street in Vancouver with the largest density of trees is W 13 Ave. For question 2, I also found that the species with the largest diameter was Japonica and there were various other species that had the largest height and smaller height/diameter.

However,  for research question 3 it remains unclear if street side really has any effect on the growth of trees. For research question 4 it is also ambiguous if tree height is a variable that should only be investigated for optimal growth as there are multiple trees with the larges heights that are scattered all throughout Vancouver. 

My research questions can be refined where research question 3 could potentially investigate the correlation between date of trees planted and the growth. Research question 4 can also be updated to looking at mapping of other characteristics of the trees throughout Vancouver. 

Further, research question 4 is revealing interesting results as the trees with the largest height are not all condensed in one location but rather spread out. 
<!----------------------------------------------------------------------------->

# Task 2: Tidy your data 

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

-   Each row is an **observation**
-   Each column is a **variable**
-   Each cell is a **value**

### 2.1 (2 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have \>8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->
```{r}
#Printing dataset for observation
print(vancouver_trees)
```
The Vancouver trees data contains tidy and untidy data: 

1. The tree_id variable is tidy as each cell has a value 
2. The civic_number is tidy as each cell has a value  
3. The cultivar_names untidy as there are missing values
4. The common_name variable is tidy as each cell has a value
5. The plant_area variable is untidy as there are missing values
6. The date_planted variable is untidy as there are missing values
7. The longitude variable is untidy as there are missing values
8. The latitude variable is untidy as there are missing values

<!----------------------------------------------------------------------------->

### 2.2 (4 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->
```{r}
#Part A: Tidy data to Untidy Data 
tidy_untidy <- vancouver_trees %>% 
  pivot_longer(cols = c("genus_name", "species_name"), 
               names_to  = "Name", 
               values_to = "Genus and Species")
print(tidy_untidy)

#Part A: Untidy data Back to Tidy Data 
untidy_original <-tidy_untidy %>% 
  pivot_wider( names_from  = "Name", 
               values_from = "Genus and Species")

print(untidy_original)

#Part B: Untidy Data to Tidy Data 
untidy_tidy <- vancouver_trees %>%  drop_na()

print(untidy_tidy)

#Part B:Tidy Data Back to Untidy Data 
tidy_original <- vancouver_trees %>% full_join(untidy_tidy)

print(tidy_original)
```
For part A I first turned tidy data into untidy data by combining the genus and species name into a new category called "Genus and Species". I then untided the data back to tidy data by using the pivot wider function. For part B I converted untidy data to tidy data by removing all NA values in the data set. I made the tidy data back to untidy data by joining the Vancouver trees data with the untidy_tidy data. 


<!----------------------------------------------------------------------------->

### 2.3 (4 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in the remaining tasks:

<!-------------------------- Start your work below ---------------------------->

1.  Does the street side name have effect on the growth of the trees in terms of height and species, or is it negligible? 
2.  Are certain species that are generally tall greatly influenced by the neighbourhood environment they are in or is their growth constant regardless of the condition? 

<!----------------------------------------------------------------------------->

Explain your decision for choosing the above two research questions.

<!--------------------------- Start your work below --------------------------->

For the first research question based on the analyses I have done above I believe that street side name does not have a major effect on  the growth of the trees and would like to further investigate the effects of date_planted with tree growth. 

For the second research question, mapping of only the maximum tree height was not significant as there were multiple large trees throughout Vancouver. The research question can be further updated in looking at tree densities, which will help visualize neighborhoods with good environments for tree growth.


<!----------------------------------------------------------------------------->

Now, try to choose a version of your data that you think will be appropriate to answer these 2 questions. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelevant columns, etc.).

(If it makes more sense, then you can make/pick two versions of your data, one for each research question.) 

<!--------------------------- Start your work below --------------------------->

# Task 3: Modelling

## 3.0 (no points)

Pick a research question from 1.2, and pick a variable of interest (we'll call it "Y") that's relevant to the research question. Indicate these.

<!-------------------------- Start your work below ---------------------------->

**Research Question**:  Are certain species that are generally tall greatly influenced by the neighbourhood environment they are in or is their growth constant regardless of the condition?

**Variable of interest**: latitude

<!----------------------------------------------------------------------------->

## 3.1 (3 points)

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen. We'll omit having to justify your choice, because we don't expect you to know about model specifics in STAT 545.

-   **Note**: It's OK if you don't know how these models/tests work. Here are some examples of things you can do here, but the sky's the limit.

    -   You could fit a model that makes predictions on Y using another variable, by using the `lm()` function.
    -   You could test whether the mean of Y equals 0 using `t.test()`, or maybe the mean across two groups are different using `t.test()`, or maybe the mean across multiple groups are different using `anova()` (you may have to pivot your data for the latter two).
    -   You could use `lm()` to test for significance of regression coefficients.

<!-------------------------- Start your work below ---------------------------->


```{r}
#Fitting model on latitude based on tree id 
fit_model <- lm(formula = latitude ~ tree_id, data = vancouver_trees)

print(fit_model)
```
<!----------------------------------------------------------------------------->

## 3.2 (3 points)

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value.

-   Be sure to indicate in writing what you chose to produce.
-   Your code should either output a tibble (in which case you should indicate the column that contains the thing you're looking for), or the thing you're looking for itself.
-   Obtain your results using the `broom` package if possible. If your model is not compatible with the broom function you're needing, then you can obtain your results by some other means, but first indicate which broom function is not compatible.

<!-------------------------- Start your work below ---------------------------->
```{r}
#Determining P value using Broom Package
pval_model <- glance(fit_model)
print(pval_model) 
```
I chose to produce a table of the statistical summaries about the model. The output tibble includes r squared, p value and other values. The pvalue is determine as 8.63 e-10. The glance function used is part of the broom package as well. 


<!----------------------------------------------------------------------------->

# Task 4: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there.

## 4.1 (3 points)

Take a summary table that you made from Task 1, and write it as a csv file in your `output` folder. Use the `here::here()` function.

-   **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
-   **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file.

<!-------------------------- Start your work below ---------------------------->
```{r}
library(readr)
write_csv(street_height_sum, "exported_file.csv")
here::here()
```

<!----------------------------------------------------------------------------->

## 4.2 (3 points)

Write your model object from Task 3 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`.

-   The same robustness and reproducibility criteria as in 4.1 apply here.

<!-------------------------- Start your work below ---------------------------->
```{r}
saveRDS(fit_model, "exported_file.rds")

readRDS(file="exported_file.rds")

```

<!----------------------------------------------------------------------------->

# Overall Reproducibility/Cleanliness/Coherence Checklist 

Here are the criteria we're looking for.

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. 

The README file should still satisfy the criteria from the last milestone, i.e. it has been updated to match the changes to the repository made in this milestone. 

## File and folder structure (1 points)

You should have at least three folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else.

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (1 point)

All output is recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant.

## Tagged release (0.5 point)

You've tagged a release for Milestone 2. 

### Attribution

Thanks to Victor Yuan for mostly putting this together.